---
title: 'Analysis report : Analysis of the m3243 cohort patient'
author: "Axel Bonesteve"
date: "2025-02-19"
output: html_document
---

```{r setup, echo=FALSE}
knitr::opts_chunk$set(message=FALSE)
knitr::opts_chunk$set(eval = TRUE)
knitr::opts_chunk$set(echo = TRUE)
```


This report is an example of the analysis performed on VCF data from patients with the human mitochondrial mutation 3243A>G, a mutation that results in leucine tRNA malformation and leads to MIDD and MELAS syndrome.

# Librairies importation

```{r}
library(data.table)
library(tidyverse)
library(paletteer)
library(ggbeeswarm)
library(BuenColors) # devtools::install_github("caleblareau/BuenColors")
library(ggpubr)
library(ggrepel)
```


# Functions 

These functions that follow were retrieved from https://github.com/caleblareau/7076.

And were initially developed for this paper : 

**Codon affinity in mitochondrial DNA shapes evolutionary and somatic fitness “**.
**Caleb A. Lareau et al. https://www.biorxiv.org/content/10.1101/2023.04.23.537997v1**

They annotate, count and statisticize synonymous variants in the mitochondrial genome based on the reference tRNA and mutated tRNA pair involved following mutation of mitochondrial genetic information. 


```{R}

annotate_synonymous <- function(df){
  sdf <- df %>% mutate(fp = Consequence == "synonymous_variant")
  variant_causes_Wobble_to_WCF <-  (grepl("\\*",sdf$Variant.tRNA) & grepl("near",sdf$Reference.tRNA)) & sdf$fp
  variant_causes_WCF_to_Wobble <-  (grepl("\\*",sdf$Reference.tRNA) & grepl("near",sdf$Variant.tRNA)) & sdf$fp
  variant_causes_Wobble_to_Wobble <-  (grepl("near",sdf$Variant.tRNA) & grepl("near",sdf$Reference.tRNA)) & sdf$fp
  variant_causes_WCF_to_WCF <- (grepl("\\*",sdf$Variant.tRNA) & grepl("\\*",sdf$Reference.tRNA)) & sdf$fp
  
  # Sanity check
  summary(as.numeric(variant_causes_Wobble_to_WCF) + 
            as.numeric(variant_causes_WCF_to_Wobble) + 
            as.numeric(variant_causes_Wobble_to_Wobble) + as.numeric(variant_causes_WCF_to_WCF))
  vec_annotate <- case_when(
    variant_causes_Wobble_to_WCF ~ "Wobble_to_WCF", 
    variant_causes_WCF_to_Wobble ~ "WCF_to_Wobble", 
    variant_causes_Wobble_to_Wobble ~ "Wobble_to_Wobble", 
    variant_causes_WCF_to_WCF ~ "WCF_to_WCF", 
    TRUE ~ "other"
  )
}

```

```{R}

count_me <- function(df){
  df %>% 
    filter(Consequence == "synonymous_variant") %>%
    group_by(syn_annotation) %>%
    summarize(n = n()) %>%
    tidyr::complete(syn_annotation, fill = list(count = 0)) %>%
    mutate(perc = n/sum(n)*100)
}

```


```{r}

process_stats_syn_type <- function(df,df_ref,count_df=FALSE, count_df_ref=FALSE) {
  
  if (count_df_ref==FALSE) {
      cm_a <- count_me(df_ref)
  } else {
      cm_a <- count_me_occ(df_ref)
  }
  if (count_df==FALSE) {  
      cm_t <- count_me(df)
  } else {     
      cm_t <- count_me_occ(df)
  }
  
  cm_t$ref_n <- cm_a$n
  cm_t$ref_perc <- cm_a$perc
  cm_t$pvalue <- c(prop.test(cm_t$n[1], sum(cm_t$n), p = cm_a$perc[1] / 100)$p.value,
                   prop.test(cm_t$n[2], sum(cm_t$n), p = cm_a$perc[2] / 100)$p.value,
                   prop.test(cm_t$n[3], sum(cm_t$n), p = cm_a$perc[3] / 100)$p.value,
                   prop.test(cm_t$n[4], sum(cm_t$n), p = cm_a$perc[4] / 100)$p.value)
  
  cm_t
}

```


# Data importation 

## m.3243A>G variants info 

```{r}

m3243_var <- read.csv2("../data/concatenate_variants.csv", header = TRUE) 

m3243_var$pos <- as.integer(m3243_var$pos) # Important step.
m3243_var$id <- paste0(m3243_var$ref, m3243_var$pos, m3243_var$alt) # Adding mutation id

# filter application second check
m3243_var$heteroplasmy_rate <- as.numeric(m3243_var$heteroplasmy_rate)
m3243_var <- m3243_var %>% filter(heteroplasmy_rate > 5)
```

### addition of majority haplogroup 

```{r}

sample.info <- m3243_var[, c("patient_id", "haplogroup")]
sample.info$hap_maj <- substr(sample.info$haplogroup, 1, 2)

sample.info.hv <- subset(sample.info, hap_maj == "HV")
sample.info.l <- subset(sample.info, hap_maj %in% paste0("L", 0:6))
sample.info.nothvandL <- subset(sample.info, !(hap_maj %in% c("HV", paste0("L", 0:6))))

sample.info.nothvandL$hap_maj <- substr(sample.info.nothvandL$hap_maj, 1, 1)
sample.info <- unique(rbind(sample.info.hv, sample.info.l, sample.info.nothvandL))

# Haplogroup join

m3243_var <- m3243_var %>%
  left_join(sample.info[,c(1,3)], by = "patient_id")

```

### addition of minority haplogroup

```{r}

sample.info = m3243_var[,c("patient_id","haplogroup")]
haplogroup = m3243_var$haplogroup
haplogroup= substr(haplogroup, start = 1, stop = 2) # Here in pairs to recover HV and L0, L1, etc.
sample.info <- data.frame(sample.info, hap_min=haplogroup)

sample.info = unique(sample.info)


# Haplogroup join
m3243_var <- m3243_var %>%
  left_join(sample.info[,c(1,3)], by = "patient_id")

```

## Random cohorte

```{r}

random_cohorte <- read.csv("../data/random_dataset_var_20250219.csv", header=TRUE)


random_cohorte$id <- paste0(random_cohorte$ref, random_cohorte$pos, random_cohorte$alt) # Adding mutation id

```

Patient information was generated by :  "MitoMatcherAnalysis/Gen_random_variant_cohorte_info/random_gen_of_m3243_var_cohorte.py" 

The information does not reflect the biological reality of a real patient.

This cohort can be replaced by a true control cohort with a new data set like this one. Or integrated into the m3243 dataset, but with another column indicating whether it is a control or a m3243 patient.


### addition of majority haplogroup 

```{r}

sample.info <- random_cohorte[, c("patient_id", "haplogroup")]
sample.info$hap_maj <- substr(sample.info$haplogroup, 1, 2)

sample.info.hv <- subset(sample.info, hap_maj == "HV")
sample.info.l <- subset(sample.info, hap_maj %in% paste0("L", 0:6))
sample.info.nothvandL <- subset(sample.info, !(hap_maj %in% c("HV", paste0("L", 0:6))))

sample.info.nothvandL$hap_maj <- substr(sample.info.nothvandL$hap_maj, 1, 1)
sample.info <- unique(rbind(sample.info.hv, sample.info.l, sample.info.nothvandL))

# Haplogroup join

random_cohorte <- random_cohorte %>%
  left_join(sample.info[,c(1,3)], by = "patient_id")

```

### addition of minority haplogroup

```{r}

sample.info = random_cohorte[,c("patient_id","haplogroup")]
haplogroup = random_cohorte$haplogroup
haplogroup= substr(haplogroup, start = 1, stop = 2) # Here in pairs to recover HV and L0, L1, etc.
sample.info <- data.frame(sample.info, hap_min=haplogroup)

sample.info = unique(sample.info)


# Haplogroup join
random_cohorte <- random_cohorte %>%
  left_join(sample.info[,c(1,3)], by = "patient_id")

```

## Annotation data

The annotation table provides by the Stanford team with a control or point of comparison. The annotation table is available in: “../data/functional_variant_tRNA_anticodon_table.tsv”.

The creation of this table is described on the github: https://github.com/EDePasquale/Mitochondrial_variants


**ref article** 
We utilized all possible mtDNA variations and protein-coding annotations as previously described by Miller et al, 2022. 
https://www.ncbi.nlm.nih.gov/pmc/articles/PMC9288977/. 
Miller TE, Lareau CA, Verga JA, DePasquale EAK, Liu V, Ssozi D, Sandor K, Yin Y, Ludwig LS, El Farran CA, Morgan DM, Satpathy AT, Griffin GK, Lane AA, Love JC, Bernstein BE, Sankaran VG, van Galen P. Mitochondrial variant enrichment from high-throughput single-cell RNA sequencing resolves clonal populations. Nat Biotechnol. 2022 Jul;40(7):1030-1034. doi: 10.1038/s41587-022-01210-8. Epub 2022 Feb 24. PMID: 35210612; PMCID: PMC9288977.

A quick explanation of how this table was created: 
they started from the NC_012920 sequence, which is the revised Cambridge sequence for human mitochondrial DNA. Against this reference, they generated the set of possible variants (single substitution, without insertion or deletion). They used this table as input for the VEP algorithm. They also queried the Mitomap database for homoplasmy, heteroplasmy, diseases associated with the variants, status, Genebank count and frequence.

**ref article** Using this landscape of 8,284 synonymous mutations, we annotated whether the codon in the reference or the alternate allele would create a canonical (i.e., Watson-Crick-Franklin, WCF) base-pairing between the codon and anticodon or would require wobble-dependent base-pairing for translation. 


```{r}

# Import data
annotations <- fread("../data/functional_variant_tRNA_anticodon_table.tsv") %>% data.frame()
annotations$mutation <- paste0("m", annotations$Position, annotations$Reference, ">", annotations$Variant) # Créer un identifiant pour chaque variant
annotations$syn_annotation <- annotate_synonymous(annotations) # On donne à chaque variant une classe.
annotations$id <- paste0(annotations$Reference,annotations$Position, annotations$Variant) # Création d'un autre identifiant.

dim(annotations)
head(annotations)
```

## Loading Mitochondrial genome loci informations


```{r}
genome_loci <- read.csv2(file = "../data/genome_loci_table.csv", header =TRUE)[-1]
head(genome_loci)
```

This dataset corresponds to all matches between loci and positions in the mitochondrial genome.


### Merging annotations and genome_loci data

Integrate annotation information into the genome_loci array, since the annotations array contains information only for variants that introduce protein modifications.


```{r}

genome_merged <- left_join(genome_loci, annotations %>% dplyr::select(-Position, Reference, Variant), by = "id") # combination of genome_loci & annotations

```

### Merging genome_merged to m3243_var

```{r}

m3243_merged <- merge(m3243_var, genome_merged, by = "id", all.x=TRUE,sort=FALSE)
colnames(m3243_merged)

m3243_merged <- m3243_merged %>% dplyr::select(-chr, -ref, -alt, -Chr, -Position, -Ref, -Alt)

```

### Merging genome_merged to random_cohorte

```{r}

random_cohorte_merged <- merge(random_cohorte, genome_merged, by = "id", all.x=TRUE,sort=FALSE)
colnames(random_cohorte)

#random_cohorte_merged <- random_cohorte_merged %>% dplyr::select(-chr,-ref, -alt, -Chr, -Position, -Ref, -Alt)

```




# Analyses

## Number of patient

The dataset contains : 

```{R}

length(unique(m3243_var$patient_id))
nrow(m3243_var %>% dplyr::select(patient_id, tissue) %>% unique()) 

```

`r length(unique(m3243_var$patient_id))` different patient IDs on `r nrow(m3243_var %>% dplyr::select(patient_id, tissue) %>% unique())` samples (id, tissue). We therefore have `r nrow(m3243_var %>% dplyr::select(patient_id, tissue) %>% unique()) - length(unique(m3243_var$patient_id))` samples belonging to an individual who therefore has several samples.



```{r}

m3243_var %>% dplyr::select(patient_id, tissue) %>% unique()

```

The array contains `r nrow(m3243_var %>% dplyr::select(patient_id, tissue) %>% unique())` rows

```{r}

length(unique(m3243_var$tissue)) # Type of tissue

```


## Distribution of mutation types

What types of mutations are there?
On which genes are they found? 
How common are these mutations?
How many silent mutations are there? 


Only one mutation per individual should be taken into account, because two samples will have the same variants for the same individual.

```{r}

m3243_merged %>% 
  dplyr::select(patient_id, id, Consequence) %>%
  unique() %>%
  group_by(Consequence) %>% 
  dplyr::count() %>% 
  arrange(desc(n)) %>% 
  mutate(cohorte = "m3243") %>%
  rbind(random_cohorte_merged %>%
          dplyr::select(patient_id, id, Consequence) %>%
          unique() %>%
          group_by(Consequence) %>%
          dplyr::count() %>%
          arrange(desc(n)) %>%
          mutate(cohorte = "controle")) %>%
  ggplot(aes(x = Consequence, y = n, group = cohorte, fill = Consequence)) +
  geom_col() + 
  geom_text(aes(label = n), position = position_stack(vjust = 0.5)) +
  facet_grid(.~cohorte) + 
  theme_minimal()+
  ylab("Nombre de variants")+
  xlab("Type de variant")+
  theme(axis.title = element_text(size = 15), axis.text = element_text(size = 13), axis.text.x = element_text(angle = 45, hjust = 1), strip.text = element_text(size = 15))

```
**Note** : NA corresponds to variants not in coding regions like tRNA, D-loop variants, rRNA, intergenic.

Is there an effect of cohort size? 

```{r}
m3243_type_var <- m3243_merged %>% 
  group_by(Consequence) %>% 
  dplyr::count() %>% 
  arrange(desc(n)) %>% 
  mutate(cohorte = "m3243")

nrow(m3243_var %>% dplyr::select(patient_id, tissue) %>% unique()) # Nb d'échantillons différents

m3243_type_var$norm <- m3243_type_var$n/ nrow(m3243_var %>% dplyr::select(patient_id, tissue) %>% unique())

random_cohorte_type_var <- random_cohorte_merged %>%
  group_by(Consequence) %>%
  dplyr::count() %>%
  arrange(desc(n)) %>%
  mutate(cohorte = "controle")

random_cohorte_type_var$norm <- random_cohorte_type_var$n/nrow(random_cohorte_merged %>% dplyr::select(patient_id, tissue) %>% unique()) 

type_var_norm <- rbind(random_cohorte_type_var, m3243_type_var)

ggplot(type_var_norm, aes(x = Consequence, y = norm, group = cohorte, fill = Consequence)) +
  geom_col() + 
  geom_text(aes(label = round(norm)), position = position_stack(vjust = 0.5)) +
  facet_grid(.~cohorte) + 
  theme_minimal()+
  ylab("Nombre de variants")+
  xlab("Type de variant")+
  theme(axis.title = element_text(size = 15), axis.text = element_text(size = 13), axis.text.x = element_text(angle = 45, hjust = 1), strip.text = element_text(size = 15))
  
  
```

### Focus on NA - intergenic, Dloop variants, tRNA, rRNA

Where are these NA variants located ? 

Cohorte m3243
```{r}

test <- subset(m3243_merged, is.na(Consequence))

pos_count_df <- test %>% dplyr::select(mutation, pos) %>% unique() %>% group_by(pos) %>% dplyr::count()

df <- data.frame(pos = 1:16569, angle = seq(0, (2*pi), by = (2*pi)/(16569-1)))

mmdf <- merge(df, pos_count_df, by.x = "pos", by.y = 'pos')


ggplot(mmdf, aes(x = n, y = pos))+
  geom_quasirandom(size = 0.1) + coord_polar(theta = "y") +
  xlim(-5,5) + ylim(16569, 1) +
  scale_color_manual(values = jdb_palette("corona"))+
  theme_void()+
  xlab("Position")+
  ylab("Mutation")+
  ggtitle("Position of tRNA, rRNA and D-loop mutations in the m3243 cohort")

```

Random cohorte

```{r}
test <- subset(random_cohorte_merged, is.na(Consequence))

pos_count_df <- test %>% dplyr::select(pos) %>% unique() %>% group_by(pos) %>% dplyr::count()

df <- data.frame(pos = 1:16569, angle = seq(0, (2*pi), by = (2*pi)/(16569-1)))

mmdf <- merge(df, pos_count_df, by.x = "pos", by.y = 'pos')


ggplot(mmdf, aes(x = n, y = pos))+
  geom_quasirandom(size = 0.1) + coord_polar(theta = "y") +
  xlim(-5,5) + ylim(16569, 1) +
  scale_color_manual(values = jdb_palette("corona"))+
  theme_void()+
  xlab("Position")+
  ylab("Mutation")+
  ggtitle("Positions of tRNA, rRNA and D-loop from random cohorte.")

```

**Note**: This representation is the position of NA variants, non-protein-coding variants, i.e. variants for the D-loop, tRNA, rRNA and intergenic region, but we observe here for the random cohort that we have these variants everywhere. This is due to the generation of variants which has no biological significance as it is simply a random choice between 4 nucleotides. 

**Upgrade**: To improve randomness, we should use the list of all possible mutations generated by Lareau and choose randomly from this list.  

What about a complete count ? 

Taking into account not only the distribution of the variants in the mtgenome but also the distribution inside the cohort.


```{r}
test <- subset(m3243_merged, is.na(Consequence))

pos_count_df <- test %>% dplyr::select(patient_id, pos) %>% unique() %>% group_by(pos) %>% dplyr::count()

df <- data.frame(pos = 1:16569, angle = seq(0, (2*pi), by = (2*pi)/(16569-1)))

mmdf <- merge(df, pos_count_df, by.x = "pos", by.y = 'pos')


m3243_NA <- ggplot(mmdf, aes(x = n, y = pos))+
  geom_quasirandom(size = 0.1) + coord_polar(theta = "y") +
  xlim(-5,10) + ylim(16569, 1) +
  scale_color_manual(values = jdb_palette("corona"))+
  theme_minimal()+
  xlab("Position")+
  ylab("Mutation")+
  ggtitle("Position mutations NA dans la cohorte m3243 ")

m3243_NA70 <- ggplot(mmdf, aes(x = n, y = pos))+
  geom_quasirandom(size = 0.1) + coord_polar(theta = "y") +
  xlim(-5,70) + ylim(16569, 1) +
  scale_color_manual(values = jdb_palette("corona"))+
  theme_minimal()+
  xlab("Position")+
  ylab("Mutation")+
  ggtitle("Position of NA mutations in the m3243 cohort limited to 70")

m3243_NA10 <- ggplot(mmdf, aes(x = n, y = pos))+
  geom_quasirandom(size = 0.1) + coord_polar(theta = "y") +
  xlim(0,10) + ylim(16569, 1) +
  scale_color_manual(values = jdb_palette("corona"))+
  theme_minimal()+
  xlab("Position")+
  ylab("Mutation")+
  ggtitle("Position of NA mutations in the m3243 cohort limited to 70")

ggarrange(m3243_NA70, m3243_NA10)

```



Random cohorte (or control cohorte)

```{r}

test <- subset(random_cohorte_merged, is.na(Consequence))

pos_count_df <- test %>% dplyr::select(patient_id, pos) %>% unique() %>% group_by(pos) %>% dplyr::count()

df <- data.frame(pos = 1:16569, angle = seq(0, (2*pi), by = (2*pi)/(16569-1)))

mmdf <- merge(df, pos_count_df, by.x = "pos", by.y = 'pos')


control_NA <- ggplot(mmdf, aes(x = n, y = pos))+
  geom_quasirandom(size = 0.1) + coord_polar(theta = "y") +
  xlim(-5,10) + ylim(16569, 1) +
  scale_color_manual(values = jdb_palette("corona"))+
  theme_minimal()+
  xlab("Position")+
  ylab("Mutation")+
  ggtitle("Position des mutations NA dans la cohorte contrôle")
```

```{r}
ggarrange(control_NA, m3243_NA)
```

**Note** : as said before, perfect random distribution for random cohort due to unrealistic mt variants generation.


### Focus on stop_retained and stop_lost variants


```{r}

m3243_merged %>% filter(Consequence == "stop_retained_variant")

random_cohorte_merged %>% filter(Consequence == "stop_retained_variant")
```

```{r}

m3243_merged %>% filter(Consequence == "stop_lost")

random_cohorte_merged %>% filter(Consequence == "stop_lost")

```


### Focus on missense variants


```{r}
missense_m3243 <- m3243_merged %>% filter(Consequence == "missense_variant")

missense_random <- random_cohorte_merged %>% filter(Consequence == "missense_variant")

```

```{r}
dim(missense_m3243)

dim(missense_random)
```


```{r}

missense_m3243 %>% dplyr::select(mutation) %>% unique()

missense_random %>% dplyr::select(mutation) %>% unique()
```

What is the distribution of these missense variants in the mitochondrial genome?

random cohort
```{r}

pos_count_df <- missense_random %>% dplyr::select(patient_id, pos) %>% unique() %>% group_by(pos) %>% dplyr::count()

df <- data.frame(pos = 1:16569, angle = seq(0, (2*pi), by = (2*pi)/(16569-1)))

mmdf <- merge(df, pos_count_df, by.x = "pos", by.y = 'pos')

random_missense_plot <- ggplot(mmdf, aes(x = n, y = pos))+
  geom_quasirandom(size = 0.1) + coord_polar(theta = "y") +
  xlim(-5,70) + ylim(16569, 1) +
  scale_color_manual(values = jdb_palette("corona"))+
  theme_minimal()+
  xlab("Position")+
  ylab("Mutation")+
  ggtitle("Position of missense mutations in the control cohort")

```


Cohorte m3243
```{r}

pos_count_df <- missense_m3243 %>% dplyr::select(patient_id, pos) %>% unique() %>% group_by(pos) %>% dplyr::count()

df <- data.frame(pos = 1:16569, angle = seq(0, (2*pi), by = (2*pi)/(16569-1)))

mmdf <- merge(df, pos_count_df, by.x = "pos", by.y = 'pos')

m3243_missense_plot <- ggplot(mmdf, aes(x = n, y = pos))+
  geom_quasirandom(size = 0.1) + coord_polar(theta = "y") +
  xlim(-5,70) + ylim(16569, 1) +
  scale_color_manual(values = jdb_palette("corona"))+
  theme_minimal()+
  xlab("Position")+
  ylab("Mutation")+
  ggtitle("Position of missense mutations in the m3243 cohort")

```

```{r}

ggarrange(random_missense_plot,m3243_missense_plot)

```

```{r}

missense_m3243 %>% 
  dplyr::select(patient_id, pos) %>% 
  unique() %>% group_by(pos) %>% 
  dplyr::count() %>% mutate(cohorte = "m3243") %>% 
  arrange(desc(n)) %>%
  rbind(missense_random %>% 
          dplyr::select(patient_id, pos) %>% 
          unique() %>% 
          group_by(pos) %>% 
          dplyr::count() %>% mutate(cohorte = "random") %>%
          arrange(desc(n))) %>%
  ggplot(aes(x = cohorte, y = n)) + 
  geom_violin()+
  theme_minimal()+
  ylab("Comptage du nombre de mutations") + 
  theme(axis.title = element_text(size = 15), axis.text = element_text(size = 13), axis.text.x = element_text(angle = 45, hjust = 1), strip.text = element_text(size = 15))


```

And now by gene? 

Import mitochondrial gene size

```{r}
mt_gene_size <- fread("../data/mtGeneSize.txt")
```


```{r}
mt_gene_size <- mt_gene_size %>% rename(Symbol = proteins, gene_size = n) 

gene_count_missense_m3243 <- missense_m3243 %>% 
  dplyr::select(patient_id, pos, Symbol) %>% 
  unique() %>% group_by(Symbol) %>% 
  dplyr::count() %>% mutate(cohorte = "m3243") %>% 
  arrange(desc(n)) %>% left_join(mt_gene_size, by = "Symbol")

nb_ind_m3243 <- length(unique(m3243_var$patient_id))
nb_ind_random <- length(unique(random_cohorte$patient_id))
  
# Normalisation par la taille des gènes et aussi par la taille de la cohorte.
gene_count_missense_m3243$prop <- gene_count_missense_m3243$n / (gene_count_missense_m3243$gene_size * nb_ind_m3243)


gene_count_missense_control <- missense_random %>% 
  dplyr::select(patient_id, pos, Symbol) %>% 
  unique() %>% group_by(Symbol) %>% 
  dplyr::count() %>% mutate(cohorte = "random") %>% 
  arrange(desc(n)) %>% left_join(mt_gene_size, by = "Symbol")

# Normalisation par la taille des gènes et aussi par la taille de la cohorte.
gene_count_missense_control$prop <- gene_count_missense_control$n / (gene_count_missense_control$gene_size * nb_ind_random)

gene_count_cohortes <- rbind(gene_count_missense_control, gene_count_missense_m3243)


ggplot(gene_count_cohortes, aes(x = Symbol, y = prop, fill = cohorte)) +
  geom_bar(stat="identity")+ 
  facet_grid(.~cohorte) + 
  theme_minimal() + 
  ylab("Valeurs de comptage de mutation normalisées par 
       la taille de gène et de cohorte") +
  theme(axis.title = element_text(size = 15), axis.text = element_text(size = 10), axis.text.x = element_text(angle = 45, hjust = 1), strip.text = element_text(size = 15))

```


Analysis of missense variants :


The rest of the analysis on these missense (pathogenic) variants is written in the Rmarkdown file named pathos_missense_variant.Rmd. Loading of m3243 variant data, from the search for missense (pathogenic) variants reported as such in the literature, found in the m3243 cohort. Annotation data was also loaded and merged with variant data. The aim is to report on the summarized clinical picture the pathogenic variants carried by patients.

First, save the lists of missense variants by cohort:

```{r}

m3243_merged %>% 
  filter(Consequence == "missense_variant") %>% 
  dplyr::select(id) %>% 
  unique() %>% 
  pull(id)  #%>% 
  #write(file = "../transformed_data/24-07-24_data/missense_variants_m3243.txt", sep = ";")


random_cohorte_merged %>%
  filter(Consequence == "missense_variant") %>%
  dplyr::select(id) %>%
  unique() %>%
  pull(id) #%>%
  #write(file = "../transformed_data/24-07-24_data/missense_variants_control.txt", sep = ";")

```

Then run the script to identify them.

The script used to search for missense variants :

```{python, eval = FALSE}

import sys
import csv
import os

# Vérifier si les fichiers d'entrée et de sortie sont spécifiés
if len(sys.argv) != 4:
    print("Utilisation : python script.py <fichier_variants.txt> <fichier_recherche_pathos.csv> <output.txt>")
    sys.exit(1)

input_variants = sys.argv[1]
search_file = sys.argv[2]
output_file = sys.argv[3]

# Vérifier si les fichiers d'entrée existent
if not all(map(lambda x: os.path.isfile(x), [input_variants, search_file])):
    print("Un ou plusieurs fichiers d'entrée sont manquants.")
    sys.exit(1)

# Ouvrir le fichier de sortie en mode écriture
with open(output_file, 'w') as outfile:
    # Lire les variants à partir du fichier d'entrée
    with open(input_variants, 'r') as infile:
        for line in infile:
            # Extraire le variant de la ligne courante
            variant = line.strip().split(';')[0]

            # Rechercher la présence du variant dans le fichier de recherche
            with open(search_file, 'r') as searchfile:
                reader = csv.reader(searchfile, delimiter=';')
                found = False
                for row in reader:
                    if row[3] == variant:
                        outfile.write(';'.join(row) + '\n')
                        found = True
                        break

                # Écrire "absent" dans le fichier de sortie si le variant n'est pas trouvé
                # if not found:
                #     outfile.write("absent\n")

print(f"Le résultat de la recherche a été enregistré dans {output_file}")

```

# Session info

```{r}
sessionInfo()
```









