---
title: 'Analysis report : Mitochondrial tRNA analysis of patients with the 3243 mutation'
author: "Axel Bonesteve"
date: "2025-02-19"
output: html_document
---

```{r setup, echo=FALSE}
knitr::opts_chunk$set(message=FALSE)
knitr::opts_chunk$set(eval = TRUE)
knitr::opts_chunk$set(echo = TRUE)
```

This report is an example of an analysis carried out on mitochondrial tRNA in a cohort of patients with mitochondrial diseases, in particular affected by the m3243 mutation responsible for MELAS and MIDD syndromes. 

# Librairies


```{r}
library(data.table)
library(tidyverse)
library(formattable)
library(randomcoloR)
library(factoextra)
library(FactoMineR)

```

# Functions

## Importation des données

### Fonctions

```{R}

annotate_synonymous <- function(df){
  sdf <- df %>% mutate(fp = Consequence == "synonymous_variant")
  variant_causes_Wobble_to_WCF <-  (grepl("\\*",sdf$Variant.tRNA) & grepl("near",sdf$Reference.tRNA)) & sdf$fp
  variant_causes_WCF_to_Wobble <-  (grepl("\\*",sdf$Reference.tRNA) & grepl("near",sdf$Variant.tRNA)) & sdf$fp
  variant_causes_Wobble_to_Wobble <-  (grepl("near",sdf$Variant.tRNA) & grepl("near",sdf$Reference.tRNA)) & sdf$fp
  variant_causes_WCF_to_WCF <- (grepl("\\*",sdf$Variant.tRNA) & grepl("\\*",sdf$Reference.tRNA)) & sdf$fp
  
  # Sanity check
  summary(as.numeric(variant_causes_Wobble_to_WCF) + 
            as.numeric(variant_causes_WCF_to_Wobble) + 
            as.numeric(variant_causes_Wobble_to_Wobble) + as.numeric(variant_causes_WCF_to_WCF))
  vec_annotate <- case_when(
    variant_causes_Wobble_to_WCF ~ "Wobble_to_WCF", 
    variant_causes_WCF_to_Wobble ~ "WCF_to_Wobble", 
    variant_causes_Wobble_to_Wobble ~ "Wobble_to_Wobble", 
    variant_causes_WCF_to_WCF ~ "WCF_to_WCF", 
    TRUE ~ "other"
  )
}

```

# Data importation

Loading data for m3243 cohort variants

```{r}

m3243 <- read.csv2("../data/concatenate_variants.csv", header = TRUE,sep = ";") 
m3243 <- m3243[,-15]
m3243$pos <- as.integer(m3243$pos)

m3243$id <- paste0(m3243$ref, m3243$pos, m3243$alt)
m3243$heteroplasmy_rate <- as.numeric(m3243$heteroplasmy_rate)

```


### Loading genomic positions

```{r}

genome_loci <- read.csv2(file = "../data/genome_loci_table.csv", header =TRUE)[,-1]
```



### Chargement des données annotations

```{r}

# Import data
annotations <- fread("../data/functional_variant_tRNA_anticodon_table.tsv") %>% data.frame()
annotations$syn_annotation <- annotate_synonymous(annotations) # On donne à chaque variant une classe.
annotations$id <- paste0(annotations$Reference,annotations$Position, annotations$Variant) # Création d'un autre identifiant.
```

### Merging annotations and genome_loci data

Integrate annotation information into the genome_loci array, since the annotations array contains information only for variants that introduce protein modifications.


```{r}

genome_merged <- left_join(genome_loci, annotations %>% dplyr::select(-Position, Reference, Variant), by = "id") # combination of genome_loci & annotations

```


### Merging annotation data and sub-cohort variant data

Merging of genome and mutation data in m3243.

```{r}
m3243_merged <- merge(m3243, genome_merged, by = "id", all.x=TRUE,sort=FALSE)
```

## Analyses

### Types de variants

```{r}

vec_name_tRNA <- scan("../data/tRNA_names.txt",character(), quote = "", sep = ",") #Chargemetn du nom des ARNt

#  Creation of a column to define the type of variant.

m3243_merged <- m3243_merged %>% mutate(type_variant = case_when(
  Biotype == "protein_coding" ~ "protein_coding", # variants protéiques
    Locus %in% c("MT-RNR1", "MT-RNR2") ~ "RNAr_coding", # Variants ARNr
    Locus %in% vec_name_tRNA ~ "RNAt_coding", # Variants ARNt
    Locus == "MT-CR" ~ "CR_coding"# Variants Coding region
))
# Create a column for tRNA, rRNA, D-loop or protein

# Calculation of variant types 

m3243_merged %>% 
  group_by(type_variant) %>%
  dplyr::count() %>%
  ggplot(aes(x = type_variant, y = n)) + 
  geom_col(position = "dodge", stat = "identity") + 
  theme_minimal() + 
  ylab("number of variants normalized by cohort size") +
  xlab("type of variant") +
  ggtitle("Number of variants according to type.") +
  theme(axis.title = element_text(size = 10), legend.text = element_text(size = 10))
```
### ARNt

Adding APOGEE2 scores

```{r}

apogee_trna <- read.csv2("../data/apogee2_trna.csv", header = T)

apogee_trna$id <- paste0(apogee_trna$Ref, apogee_trna$Pos, apogee_trna$Alt)
```

Fusion with the other datasets and subselection of tNRA 

```{r}

m3243_trna <- m3243_merged %>% left_join(apogee_trna %>% dplyr::select(id, t.APOGEE.score, t.APOGEE.unbiased.score), by = "id") %>% filter(Locus %in% vec_name_tRNA)

```

Sous-sélection : 

```{r}
m3243_trna %>% filter(type_variant == "RNAt_coding") %>% 
  group_by(Locus) %>%
  dplyr::count() %>%
  ggplot(aes(x = Locus, y = n)) +
  geom_bar(position = "dodge", stat = "identity") +
  ylab("Number of variants")+
  xlab("tRNA variants")+
  ggtitle("Number of variants present on tRNAs m3243 cohort.")+
  theme(axis.title = element_text(size = 10), legend.text = element_text(size = 10))+
  theme_minimal()+
  scale_y_continuous(
    breaks = seq(1,10, by = 1)
  )
```

Evolution of the apogee score for tRNAs

```{r}
# absence of associated probability in t-RNA APOGEE data provided by MitImpact

# m3243_trna %>% 
#   dplyr::select((id, cohorte, t.APOGEE.score.x, t.APOGEE.unbiased.score) %>%
#   unique() %>%
#   ggplot(aes(x = APOGEE2_score, y = APOGEE2_probability, color = APOGEE2))+
#   geom_point() + 
#   theme_light() + 
#   ylab("Probabilité APOGEE2")+
#   xlab("Score APOGEE2")+
#   ggtitle("Evolution du score APOGEE2 en fonction de la probabilité.")+
#   theme(axis.title = element_text(size = 10), legend.text = element_text(size = 10), strip.text = element_text(size = 10) )+
#   facet_grid(.~cohorte)


```


```{r}

m3243_trna %>% filter(type_variant == "RNAt_coding" & Locus == "MT-TT") %>% dplyr::select(id, heteroplasmy_rate, patient_id, haplogroup, t.APOGEE.score, t.APOGEE.unbiased.score)

```


```{r}

m3243_trna %>% filter(type_variant == "RNAt_coding" & Locus == "MT-TL2") %>% dplyr::select(id, heteroplasmy_rate, patient_id, haplogroup, t.APOGEE.score, t.APOGEE.unbiased.score)

```

```{r}
sessioninfo::session_info()
```


